Monitoring of Cell-free Human Papillomavirus DNA in Metastatic or Recurrent Cervical Cancer: Clinical Significance and Treatment Implications

Version 1

Abstract

Abstract

Purpose

Monitoring of circulating human papillomavirus (HPV) cell-free DNA (cfDNA) is a minimally invasive approach for surveillance in HPV-associated cancers, particularly cervical cancer. The aim of this study was to monitor circulating HPV cfDNA levels in patients with recurrent or metastatic cervical cancer during treatment and follow-up to assess the utility of HPV cfDNA as a tumor marker for disease surveillance and in guiding clinical treatment decisions.

Experimental Design

In this prospective pilot observational study, levels of HPV cfDNA in serum samples from 28 patients with recurrent or metastatic HPV+ cervical cancer were measured via digital droplet polymerase chain reaction. Results for HPV cfDNA levels were matched to clinical outcomes and to serum levels of squamous cell carcinoma antigen (SCC-Ag) to assess the clinical potential of HPV cfDNA as a tumor marker.

Results

HPV cfDNA was detected in all 28 patients (100%). Notably, median baseline HPV cfDNA levels varied according to the metastatic pattern in individual patients (P=0.019). Specifically, patients with a combined multiple-metastasis pattern had higher median baseline HPV cfDNA levels than patients with a single metastasis (P=0.003). All participants exhibited changes in HPV cfDNA levels over a median monitoring period of 2 months (range 0.3–16.9) before evaluations for treatment response or disease progression. Among 26 patients initially diagnosed with squamous cell cervical cancer, the positivity rate was 100% for HPV cfDNA and 69.2% for SCC-Ag (P=0.004, 95% confidence interval, 0–0.391). Among 20 patients longitudinally monitored for squamous cell cervical cancer, the concordance with changes in disease status was 90% for HPV cfDNA and 50% for SCC-Ag (P=0.014, 95% confidence interval, 0.022–0.621).

Conclusions

HPV cfDNA is a promising tumor marker for HPV+ cervical cancer that offers advantages over SCC-Ag. In the context of precision medicine, HPV cfDNA is poised to play an increasingly pivotal role in monitoring treatment efficacy, providing valuable insights into disease progression, and guiding clinical decisions.

Introduction

Cervical cancer (CC) is the fourth most prevalent malignancy in terms of both incidence and mortality among females globally, and is the primary human papillomavirus (HPV)-associated cancer [1, 2]. Despite advances in CC treatment, challenges persist regarding recurrence and metastasis [3]. The treatment landscape for recurrent or metastatic CC has changed with the emergence of targeted immunotherapy drugs. However, the lack of effective biomarkers hinders assessment of treatment efficacy and the ability to predict patient outcomes in this setting [4]. There is a pressing need for more effective and minimally invasive biomarkers for serial monitoring of treatment responses and prognostication of patient outcomes.

As a liquid biopsy modality, measurement of circulating cell-free DNA (cfDNA) released from tumor cells into the bloodstream has extensive utility in optimizing various facets of cancer management, including early diagnosis [5], noninvasive genotyping, pretreatment assessment, drug target identification, resistance detection [6, 7], treatment efficacy monitoring, post-treatment follow-up, and relapse prediction [8–10]. HPV is the cause of most CC cases [11]. Viral DNA from high-risk HPV subtypes integrates into host cell genomes, resulting in widespread expression of virus-specific E6/7 protein [12]. As HPV-associated cancers release HPV cfDNA into the host’s bloodstream, circulating HPV cfDNA in free or integrated form is an attractive potential biomarker that can be detected in blood [13, 14]. Typically, circulating cfDNA is detected using digital droplet polymerase chain reaction (ddPCR) or next-generation sequencing (NGS) technology [15]. ddPCR allows direct, independent, and absolute quantification of cfDNA in samples, with a DNA detection threshold as low as 1 copy/mL. Results can be obtained within 1 day, so ddPCR is relatively cost-effective [16, 17].

We conducted a prospective clinical study (NCT03175848) in primary stage IVB CC to investigate the role of radiotherapy (RT) in combination therapy. Recent advances in targeted therapies and immunotherapies have significantly enhanced the treatment efficacy for recurrent or metastatic CC and have increased overall survival (OS) [4]. Some patients may require maintenance therapy for up to 2 years, and others may survive with their tumor for an extended period. Consequently, it is imperative to closely monitor each patient’s condition and adjust their treatment plan accordingly. However, conventional imaging methods occasionally fail to reflect disease changes in a timely manner [18], while blood biomarkers such as squamous cell carcinoma antigen (SCC-Ag) have limited clinical effectiveness for monitoring [19]. Considering these clinical challenges, we hypothesized that circulating HPV cfDNA levels correlate with metastatic patterns and treatment response in CC. To validate this hypothesis, we prospectively recruited an observational pilot cohort comprising patients with primary stage IVB or recurrent HPV-positive CC and measured HPV cfDNA levels for analysis of HPV cfDNA copy numbers in relation to disease parameters and treatment responses.

Methods

Study design

From August 2017 to February 2023, a total of 33 patients with pathologically confirmed primary stage IVB or recurrent HPV+ CC were enrolled at Zhejiang Cancer Hospital. Five patients were excluded from the analysis, as outlined in Figure 1. The final analysis cohort comprised 28 cases: 21 with primary stage IVB CC and seven with recurrent CC. Notably, 19 cases in the primary CC group participated in the prospective clinical study (NCT03175848). The main eligibility criteria were: pathologically confirmed diagnosis of CC; pathological evidence of at least one metastatic lesion; and PCR positivity for a high-risk HPV subtype in pretreatment exfoliated cervical cells or a serum sample. All patients consented to the study protocol, including collection of blood samples throughout the study. Patients received chemotherapy with or without immunotherapy, targeted therapy, or RT. For patients with primary stage IVB CC who had a single sample collected, this occurred at treatment initiation. For the group undergoing longitudinal sampling, patients with primary stage IVB CC had three to five blood samples collected at treatment initiation, mid-treatment, and during follow-up, while patients with recurrent CC had three 3 blood samples collected during treatment, starting from enrollment. For all serum specimens, HPV cfDNA was quantified using ddPCR. Concurrent SCC-Ag testing was conducted for patients with squamous cell CC in the sequential sampling group (at a time point matching or closely following HPV cfDNA sampling). The 2018 International Federation of Gynecology and Obstetrics staging criteria were applied. Each patient underwent routine imaging assessments, and treatment efficacy was evaluated according to Response Evaluation Criteria in Solid Tumors (RECIST version 1.1). We defined lymph node metastasis as metastasis in para-aortic or distant lymph nodes (e.g., supraclavicular, inguinal, mediastinal nodes). We classified patterns of CC recurrence or metastasis into five groups: local recurrence (LR); lymph node metastasis (LNM); hematogenous metastasis (HM); lymph node+hematogenous metastasis (LN+HM); and lymph node+ hematogenous+diffuse serosal metastasis (LN+H+DSM). DSM encompasses metastases to the peritoneal, pleural, or pericardial regions. The study was approved by the Ethics Committee of Zhejiang Cancer Hospital.

CONSORT flow diagram showing the enrollment of an observational cohort of patients diagnosed with recurrent or metastatic HPV-cervical cancer undergoing serum human papillomavirus (HPV) cell-free DNA surveillance.

HPV diagnostics and typing

HPV subtypes were determined via routine PCR for exfoliated cervical cells or serum samples collected before treatment. Patients were excluded from the study if both samples tested negative. A commercial PCR testing kit designed primarily for qualitative genotyping of HPV DNA extracted from exfoliated cervical cells was routinely used. This kit can detect 21 HPV virus subtypes (8 low-risk and 13 high-risk subtypes): HPV16, HPV18, HPV31, HPV33, HPV35, HPV39, HPV45, HPV51, HPV52, HPV56, HPV58, HPV59, and HPV68.

Samples

Serum was extracted from peripheral whole blood samples at the Radiobiology Laboratory of Zhejiang Cancer Hospital. A 5-mL aliquot of whole blood was collected into a yellow-top blood collection tube and allowed to clot at room temperature for 30 min. After centrifugation at 2000×g for 10 min in a refrigerated centrifuge, the serum was carefully transferred into polypropylene tubes in 1-mL aliquots and stored at −80°C. Frozen serum samples were subsequently transported to the Oncology Research Institute of Zhejiang Cancer Hospital for DNA extraction.

Procedure for ddPCR analysis

Before analysis, serum samples were thawed and centrifuged at 2000×g at 4°C for 10 min for DNA extraction. In accordance with the manufacturer’s protocol, cfDNA was isolated from 2 mL of serum using a QIAamp Circulating Nucleic Acid Extraction Kit (Qiagen, Hilden, Germany). The DNA was eluted twice through a column for purification, resulting in 60 μL of eluate that was stored at −80°C until analysis. Primers and probes for ddPCR detection were designed on the basis of E7 gene sequences of the target HPV subtypes to generate amplicons of varying length (Supplementary Table 1). Each ddPCR reaction used 30 μL of DNA template. According to the manufacturer’s instructions for the QIAcuity QX-200 ddPCR platform (Qiagen), ddPCR reactions consisted of 40 µL of reaction mixture per well that included the primers, probes, and template. The reactions were amplified in QIAcuity 26,000 24-well Nanoplates (Qiagen) under the following conditions: initial enzyme activation at 95°C for 2 min, followed by 50 cycles of denaturation at 95°C for 15 s and annealing at 60°C for 30 s. The exposure time for imaging of partitions was 400 ms for fluorescein amidite (FAM) and 300 ms for 2’-chloro-7’-phenyl-1,4-dichloro-6-carboxyfluorescein (VIC). Data analysis was performed using QIAcuity software version 2.1.7 (Qiagen) to quantify HPV copy numbers.

HPV cfDNA monitoring protocol

Participants were enrolled on a rolling basis, and the first serum sample collected was regarded as the baseline sample. A baseline sample was categorized as a treatment initiation sample if it was collected between Day −14 and Day +30 preceding initial treatment in patients with primary stage IVB CC, or before treatment for relapse or disease progression in patients with recurrent CC.

For patients with sequential samples, the time at which the initial blood sample was collected was designated as time 0. Up to four additional samples were collected at intervals ranging from 8 to 962 days (median 73 days, mean 128 days), and blood collection continued for up to 1513 days. Blood sample collection was coordinated with patient treatment and follow-up times to the greatest extent possible for measurement of HPV cfDNA levels (copies/mL). Serum SCC-Ag levels are routinely assessed multiple times before, during, and after treatment for patients with squamous cell CC at our hospital using an Abbott Architect instrument (Abbott Laboratories, Abbott Park, IL, USA). To match the analysis, we selected SCC-Ag values measured at the same time as the HPV serum sample or at the most recent time point. To ensure consistency in the analysis, we chose SCC-Ag values measured concurrently with the HPV serum sample or at the latest time point available. SCC-Ag levels <1.5 ng/mL were classified as normal, while levels ≥1.5 ng/mL; the upper limit of detection limit was 70 ng/mL. For statistical purposes, clinical test results exceeding 70 ng/mL were treated as 70 ng/mL.

Following the literature [20, 21], serum samples were deemed HPV+ if at least three droplets containing HPV amplicons were identified. Samples with fewer than three droplets containing HPV amplicons or no amplicons detected were categorized as HPV-negative. Serum HPV cfDNA levels were quantified as copies/mL.

Statistical analyses

Statistical analyses were performed using GraphPad Prism 9. The Mann-Whitney U test was used to assess the difference in the number of viral DNA copies between two groups or two metastasis patterns. Kendall’s τ correlation test was used to determine the coefficient of correlation between two factors. Comparisons of the rates were conducted using Fisher’s exact test. Kaplan-Meier survival analysis was conducted to calculate the hazard ratio with a 95% confidence interval (CI) using the Cox model. Overall survival (OS) was defined as time from diagnosis until death from any cause or the last follow-up date (December 31, 2023). All P-values reported are two-tailed, with statistical significance defined as P<0.05.

Results

Patient characteristics

Our study included 28 patients diagnosed with HPV+ advanced CC treated at our hospital, comprising 21 cases (75%) with primary stage IVB CC and seven cases (25%) with recurrence and metastasis after treatment. The clinical characteristics of the patients are listed in Table 1. The median age at diagnosis was 52 years (range 34–67). In total, 76 serum samples were obtained from the cohort, consisting of 69 longitudinal samples from 21 patients (3–5 samples per patient) and single-time samples from seven patients at treatment initiation. Twenty-five (89%) of the baseline samples were obtained at treatment initiation. HPV cfDNA was quantified in all serum samples via ddPCR.

Validity of the test for HPV cfDNA detection and genotyping

To validate the consistency of HPV genotyping between serum and exfoliated cervical cells from the same patient, we conducted conventional PCR genotyping of HPV cfDNA in baseline serum. The analysis revealed that only 12/28 patients (42.9%) tested positive; however, the genotyping outcomes were entirely consistent with those obtained from HPV genotyping in matched cervical exfoliated cells. Subsequently, the 28 baseline serum samples underwent qualitative and quantitative assessment using ddPCR. The results revealed that all 28 patients (100%) tested positive for HPV cfDNA, with HPV typing showing complete concordance (100%) with HPV PCR outcomes from matched cervical exfoliated cells. HPV genotyping revealed that HPV16, HPV58, HPV18, and other subtypes (31, 66, 33) accounted for 72.4%, 10.3%, 6.9%, and 10.3% of cases, respectively. (Figure 2A).

(A) Serum HPV cfDNA copy number for six HPV subtypes at baseline. (B) Baseline HPV cfDNA copy numbers for five recurrence/metastasis pattern subgroups. Statistical significance was determined using a two-sided Kruskal-Wallis test. (C) Relationship between HPV cfDNA viral copy number and metastatic pattern. Statistical significance was determined using a two-sided Mann-Whitney U test. All plots show the median and interquartile range on a log10 scale.

Correlation between tumor metastasis pattern and baseline HPV cfDNA

Analysis for the study cohort revealed an association between tumor metastatic pattern and baseline HPV cfDNA levels. According to their metastatic status at baseline, patients were categorized into two groups: the single-metastasis pattern (SMP) group (LR, LNM, or HM); and the multiple-metastasis pattern (MMP) group (LN+H±DSM). The baseline copy number significantly differed among the five recurrence/metastatic pattern groups (P=0.019) and tended to gradually increase with the degree of recurrence/metastasis (Figure 2B). The median baseline HPV cfDNA copy number was significantly higher in the MMP group than in the SMP group (P=0.003; Figure 2C).

HPV cfDNA as a predictor of treatment response or failure

We followed the enrolled patients for an average of 24.1 months (range 2.1–77.8). Figure 3 shows temporal changes in serum HPV cfDNA levels. In all patients with longitudinal testing, changes in HPV cfDNA levels occurred at a median of 2 months (range 0.3–16.9) before confirmation of a treatment response or disease progression via imaging assessment. Six patients with clinically progressive disease (as per RECIST) exhibited elevation of HPV cfDNA copy number before imaging-confirmed disease progression. The median time from detection of elevated plasma HPV cfDNA to imaging confirmation of disease progression was 4.2 months (range 1.9–16.9; Figure 3A1). Likewise, we observed a consistent decrease in HPV cfDNA copy numbers in 16 patients before imaging confirmed a treatment response. The median time from detection of a decrease in HPV cfDNA to imaging confirmation of disease regression was 1.2 months (range 0.3–2.8; Figure 3B1-D1).

Paired plots for serum HPV cfDNA (copies/mL) and squamous cell carcinoma antigen (SCC-Ag) (ng/mL) levels were measured longitudinally in patients with metastatic or recurrent HPV+ cervical cancer (lines of the same color for each patient). HPV cfDNA levels in serum were scaled using log10. (A1, A2) Patients whose disease progressed during treatment. (B1,B2) Selected patients who showed a response to cytotoxic chemotherapy. (C1,C2) Response to immune therapy in a patient with stage IVB cervical cancer positive for HPV subtypes 33 and 16, treated with paclitaxel + cisplatin, concurrent chemoradiotherapy, brachytherapy, and bevacizumab. (D1, D2) Selected patients with a response to radiotherapy.(E1, E2) HPV cfDNA levels were significantly higher than normal at some of the longitudinal time points in five patients, but SCC-Ag levels were within the normal range (<1.5 ng/mL) at all time points. (F1, F2) HPV cfDNA levels for all patients with longitudinal samples (n=21) and matched SCC-Ag levels for the patients with squamous cell cancer (n=20). Each colored line corresponds to one patient, except in C. Horizontal dashed lines indicate the days between plasma HPV cfDNA levels suggestive of a response or progression (rise or fall in levels) and imaging-confirmed changes in disease status. The associated number of days (D) is listed adjacent to the horizontal dashed line for comparison. It should be noted that not all HPV cfDNA data points are plotted for each patient.

For patients in whom systemic cytotoxic chemotherapy was effective, higher HPV cfDNA copy number at baseline was correlated with a more substantial decrease in copy number after chemotherapy. For two patients with high HPV cfDNA levels of 3.1×106 copies/mL and 1.7×106 copies/mL at baseline, the HPV cfDNA copy number decreased significantly following one or two cycles of paclitaxel+cisplatin (TP) chemotherapy (median 98.2%, range 96.7–99.5). The rate of decline in viral load slowed after the subsequent cycle of TP chemotherapy, with a median decline of 51.3% (range 39–63.5%). In addition, a patient with baseline HPV cfDNA of 2.8×104 copies/mL experienced a 45% reduction in copy number following one cycle of TP chemotherapy (Figure 3B1).

We observed that changes in HPV cfDNA levels may indicate a response to combined immunotherapy and targeted therapy. One patient diagnosed with primary stage IVB CC and multiple metastases (LN+H+DSM) tested positive for HPV33 and HPV16 in exfoliated cervical cells and serum. The patient received four cycles of TP and concurrent chemoradiotherapy plus brachytherapy for the pelvic-abdominal primary focus, followed by two additional TP cycles. The HPV33 viral load decreased from a pretreatment level of 1.5×107 copies/mL to 6.7×104 copies/mL after treatment; however, the HPV16 viral load increased from 7.7×102 to 1.2×104 copies/mL. Subsequent imaging evaluation after 58 days indicated the emergence of new foci in the lungs, prompting adjustments to the patient’s treatment plan. Guided by the immunohistochemical presence of PD-L1-positive cells (combined positive score of 10) in the primary cervical lesion, the patient received one cycle of TP + bevacizumab + anti-PD-1 treatment. This regimen led to a swift decrease in viral levels of both HPV subtypes after 36 days, with HPV 33 reaching undetectable levels and HPV 16 decreasing to 7.8×102 copies/mL (Figure 3C1).

Serum HPV cfDNA levels also changed in response to RT. A transient rise (20.8-fold) in HPV cfDNA copy number was observed in a patient with stage IVB CC who experienced a 2-week interruption of RT because of grade IV thrombocytopenia. Following stereotactic body RT to lung metastases, a patient with pulmonary oligometastases experienced a 29% decrease in HPV cfDNA copy number. Another patient diagnosed with stage IVB CC and pelvic bone metastases who underwent radical concurrent chemoradiotherapy targeting both primary and metastatic lesions exhibited a notable reduction in HPV cfDNA levels (Fig. 3D1).

Correlation between HPV cfDNA and SCC-Ag

There were 26 patients with squamous cell CC in the study cohort. All 26 (100%) had elevated serum HPV cfDNA at baseline, but only 18/26 patients (69.2%) had elevated SCC-Ag at baseline (P=0.004, 95% CI, 0–0.391). Among 72 serum samples from patients with squamous cell CC, the median HPV cfDNA level was 1.7×104 copies/mL (range 0–1.4×107) and the median SCC-Ag level was 2.6 ng/mL (range 0.4–70). There was no significant correlation between SCC-Ag and HPV cfDNA levels (R2=0.034, P=0.120). For patients with squamous cell CC who had longitudinal monitoring (n=20), the concordance with disease change was 90% for HPV cfDNA and 50% for SCC-Ag (P=0.014, 95% CI, 0.022–0.621). Comparison of matched serum HPV cfDNA and SCC-Ag levels for patients with squamous cell CC (Figure 3A–D) revealed that HPV cfDNA exhibited dynamic fluctuations, while serum SCC-Ag levels in the majority of patients rapidly decreased to near or below the normal range (<1.5 ng/mL) following initiation of treatment. During the course of treatment, SCC-Ag levels remained consistently within the normal range (<1.5 ng/mL) at all time points in five patients, but matched serum HPV cfDNA showed fluctuating changes above normal values at some time points.(Figure 3E1, E2).

Correlation between HPV cfDNA and survival

The 5-year OS rate for the entire cohort was 42.3%, with median OS of 52.1 months at median follow-up of 42.3 months (range: 10.2–88.5 months). As of December 31, 2023, there were 12 patient deaths and 20 disease progression events. Analysis of survival by HPV subtypes revealed that the difference in OS between the HPV16+ group and the non-HPV16+ group was not statistically significant (P=0.052; Figure 4A). Patients with primary stage IVB CC (n=21) were stratified by baseline median HPV cfDNA level using 3.9×104 copies/mL as the dichotomization threshold. The difference in OS between the groups with ≥3.9×104 copies/mL and <3.9×104 copies/mL at baseline was not statistically significant (P=0.111; Figure 4B).

(A) Overall survival for patients stratified by HPV subtype (HPV 16+ versus non-HPV16+). (B) Overall survival for patients stratified by baseline serum HPV cfDNA copy number using a cutoff of 3.9×104 copies/mL. P-values were calculated using a two-sided log-rank test.

Discussion

We conducted a prospective pilot observational study in patients with metastatic or recurrent CC to analyze ddPCR HPV cfDNA results in relation to SCC-Ag levels, clinical treatment responses, and prognosis. The study findings confirm the significant clinical potential of dynamic HPV cfDNA surveillance for CC. First, we found a correlation between baseline HPV cfDNA copy number and recurrence/metastasis patterns. Second, HPV cfDNA predicted treatment response or disease progression at an earlier time point than imaging assessments. Finally, we demonstrated the superiority of serum HPV cfDNA over SCC-Ag in monitoring metastatic or recurrent CC.

Our positivity rate for HPV cfDNA was 42.9% (12/28) when using PCR and 100% (28/28) when using ddPCR. We found 100% agreement of HPV typing results between exfoliated cervical cells and serum samples. These findings demonstrate the high sensitivity and specificity of HPV cfDNA detection via ddPCR. A meta-analysis comparing the accuracy of different methods for HPV cfDNA detection in HPV+ tumors revealed that NGS outperformed ddPCR and quantitative PCR in terms of sensitivity, while specificity remained consistent across all three methods [15]. We found that HPV cfDNA positivity in HPV+ CC correlated positively with tumor stage, tumor load, and lymph node status. Consistent with our findings, a recent study demonstrated a serum HPV cfDNA positivity rate of 100% in HPV+ metastatic CC [22]. Consequently, HPV cfDNA appears to be an ideal serum tumor marker for HPV+ metastatic or recurrent CC given the highly sensitive and specific detection methods available.

Studies have shown that HPV cfDNA levels correlate with disease stage, tumor size, tumor load, and lymph node status [23–25]. Accurate quantification of tumor load can be challenging, particularly in settings involving diffuse serosal metastases. Therefore, we classified recurrence/metastatic patterns into five categories and observed a significant difference in median baseline HPV cfDNA copy number among these (P=0.019). Patients were categorized into SMP and MMP groups according to their recurrence/metastasis status at baseline. The MMP group (LN+H±DSM) had higher median HPV cfDNA copy number at baseline than the SMP group (LR or LNM or HM; P=0.003). The recurrence/metastasis pattern appears to reflect tumor load and spread. A study by Mittelstadt et al. involving 35 patients with CC also revealed correlation between HPV cfDNA levels and tumor load and spread [24]. Patients with multiple metastases are likely to have a higher tumor load with greater shedding of HPV-containing DNA fragments, which can enter the blood circulation via several pathways, resulting in higher HPV cfDNA levels.

In our study, changes in HPV cfDNA levels frequently preceded confirmation of disease changes on imaging scans. Several studies demonstrated that changes in HPV cfDNA copy number are associated with response to therapy for HPV+ tumors [22, 24, 25]. We found that the median time from the onset of a change in HPV cfDNA copy number to imaging confirmation of a treatment response or disease progression was 2 months (range 0.3–16.9). Another study reported analogous findings in HPV-associated oropharyngeal cancer: changes in HPV cfDNA copy number were observed at a median of 16 days (range 12–38) before imaging confirmation of treatment response or disease progression in all 22 patients enrolled [26]. These findings suggest that HPV cfDNA can serve as a sensitive marker and a valuable clinical indicator. In addition, our results indicate that HPV cfDNA can serve as a tool for monitoring the effectiveness of RT. Among patients receiving effective RT, HPV cfDNA levels gradually declined throughout the treatment; however, interruptions in RT could lead to transient increases in HPV cfDNA levels. Moreover, our results demonstrate that the combination of immunotherapy and targeted therapy, along with cytotoxic chemotherapy, enhances tumor cell death and clearance of HPV cfDNA in vivo, surpassing the effects of cytotoxic chemotherapy alone. Our findings suggest that HPV cfDNA holds promise as a tool for evaluating the effectiveness of novel therapies such as immunotherapy and targeted therapy and informing subsequent maintenance treatment strategies for patients.

While SCC-Ag is acknowledged as a serum tumor marker for squamous cell CC [27–29], its clinical sensitivity and specificity for monitoring responses to therapy are low [30, 31]. We compared HPV cfDNA and corresponding SCC-Ag levels in 72 serum samples and found no correlation between the two data sets (R2=0.03, P=0.11). HPV cfDNA exhibited several advantages as a serum tumor marker in our study cohort. First, the positivity rate was 100% for HPV cfDNA versus 69.2% for SCC-Ag at baseline (P=0.004, 95% CI, 0–0.391). Second, changes in HPV cfDNA copy number showed greater concordance with disease progression in the group of 20 patients with squamous cell CC with longitudinal monitoring, with a concordance rate of 90% versus 50% for SCC-Ag (P=0.014, 95% CI, 0.022–0.621). Finally, HPV cfDNA provides more comprehensive information for dynamic monitoring in comparison to SCC-Ag, as SCC-Ag levels consistently remained within the normal range in some patients. Thus, HPV cfDNA is a more useful serum marker than SCC-Ag in patients with metastatic or recurrent CC.

The 5-year OS rate for the study cohort was 42.3% and the median OS time was 52.1 months. This good outcome was due to active systemic treatment (chemotherapy+immunotherapy) and local treatment (RT to primary and metastatic foci). Therefore, noninvasive dynamic monitoring using serum tumor markers is important for this population. Our data showed that correlation between OS and HPV subtypes (HPV16+ vs non-HPV16+) did not reach statistical significance (P=0.052). However, we observed a trend towards better prognosis for patients with HPV16+ than for patients with positivity for other HPV subtypes, and further expansion of the sample size may yield positive results. In the analysis of baseline copy number in relation to prognosis, the difference in OS between the groups with baseline

HPV cfDNA levels ≥3.9×104 and <3.9×104 copies/mL was not significant (P=0.111). Previous studies reported a positive correlation between high baseline HPV cfDNA levels and poor prognosis in oropharyngeal cancer [32–34], so further in-depth studies with larger sample sizes are needed to assess this correlation in CC.

Our study has some obvious limitations, including the small sample size and the heterogeneous sequential sampling protocol, for which both the number of blood draws and time interval during treatment varied. We are currently conducting a prospective study of stage IVB CC. Owing to the absence of literature support for HPV cfDNA sampling protocols, we designed this initial exploratory study with a small sample size to clarify the value of HPV cfDNA monitoring and explore various sampling times and intervals. Our study results suggest that monitoring of HPV cfDNA is valuable before, during, and after treatment. Assessment of HPV cfDNA levels in every chemotherapy cycle (monthly) during treatment and every 3–6 months during follow-up may be a reasonable approach.

In conclusion, our prospective study demonstrated that HPV cfDNA detection has high sensitivity and specificity for prediction of recurrence or metastasis in HPV-related CC, making it a suitable tumor marker for treatment monitoring and long-term follow-up. Studies have shown that the baseline HPV cfDNA copy number correlates with the patient’s metastatic pattern, which reflects in part the extent of tumor load and spread. As a serum tumor marker, HPV cfDNA outperforms SCC-Ag in reflecting changes in disease dynamics and provides a timely measure of response to various treatments, including RT and emerging immune and targeted therapies. These preliminary findings show the potential of HPV cfDNA in monitoring the efficacy of treatments for HPV-associated cancers and predicting disease progression and recurrence, but large-scale prospective validation in clinical trials is required.

eLife Assessment

In the field of early detection of disease and recurrence and monitoring of treatment efficacy by cfDNA analysis, this study presents a useful finding that HPV cfDNA level monitoring provides advantages over serum levels of squamous cell carcinoma antigen (SCC-Ag), specifically for HPV+ cervical cancer. The data were collected and analysed using solid and validated methodology but the sample number was limited.

Reviewer #1 (Public review):

Summary:

The study "Monitoring of Cell-free Human Papillomavirus DNA in Metastatic or Recurrent Cervical Cancer: Clinical Significance and Treatment Implications" by Zhuomin Yin and colleagues focuses on the relationship between cell-free HPV (cfHPV) DNA and metastatic or recurrent cervical cancer patients. It expands the application of cfHPV DNA in tracking disease progression and evaluating treatment response in cervical cancer patients. The study is overall well-designed, including appropriate analyses.

Strengths:

The findings provide valuable reference points for monitoring drug efficacy and guiding treatment strategies in patients with recurrent and metastatic cervical cancer. The concordance between HPV cfDNA fluctuations and changes in disease status suggests that cfDNA could play a crucial role in precision oncology, allowing for more timely interventions. As with similar studies, the authors used Droplet Digital PCR to measure cfDNA copy numbers, a technique that offers ultrasensitive nucleic acid detection and absolute quantification, lending credibility to the conclusions.

Weaknesses:

Despite including 28 clinical cases, only 7 involved recurrent cervical cancer, which may not be sufficient to support some of the authors' conclusions fully. Future studies on larger cohorts could solidify HPV cfDNA's role as a standard in the personalized treatment of recurrent cervical cancer patients.

Reviewer #2 (Public review):

Summary:

The authors conducted a study to evaluate the potential of circulating HPV cell-free DNA (cfDNA) as a biomarker for monitoring recurrent or metastatic HPV+ cervical cancer. They analyzed serum samples from 28 patients, measuring HPV cfDNA levels via digital droplet PCR and comparing these to squamous cell carcinoma antigen (SCC-Ag) levels in 26 SCC patients, while also testing the association between HPV cfDNA levels and clinical outcomes. The main hypothesis that the authors set out to test was whether circulating HPV cfDNA levels correlated with metastatic patterns and/or treatment response in HPV+ CC.

The main claims put forward by the paper are that:

(1) HPV cfDNA was detected in all 28 CC patients enrolled in the study and levels of HPV cfDNA varied over a median 2-month monitoring period.

(2) 'Median baseline' HPV cfDNA varied according to 'metastatic pattern' in individual patients.

(3) Positivity rate for HPV cfDNA was more consistent than SCC-Ag.

(4) In 20 SCC patients monitored longitudinally, concordance with changes in disease status was 90% for HPV cfDNA.

This study highlights HPV cfDNA as a promising biomarker with advantages over SCC-Ag, underscoring its potential for real-time disease surveillance and individualized treatment guidance in HPV-associated cervical cancer.

Strengths:

This study presents valuable insights into HPV+ cervical cancer with potential translational significance for management and guiding therapeutic strategies. The focus on a non-invasive approach is particularly relevant for women's cancers, and the study exemplifies the promising role of HPV cfDNA as a biomarker that could aid personalized treatment strategies.

Weaknesses:

While the authors acknowledge the study's small cohort and variability in sequential sampling protocols as a limitation, several revisions should be made to ensure that (1) the findings are presented in a way that aligns more closely with the data without overstatement and (2) that the statistical support for these findings is made more clear. Specific suggestions are outlined below.

(1) The authors should provide source data for Figures 2, 3, and 4 as supplementary material.

(2) Description of results in Figure 2: Figure 2 would benefit from clearer annotations regarding HPV virus subtypes. For example, does the color-coding in Figure 2B imply that all samples in the LR subgroup are of type HPV16? If that is the case, is it possible that detection variations are due to differences in subtype detection efficiency rather than cfDNA levels? The authors should clarify these aspects. Annotation of Figure 2B suggests that the p-value comes from comparing the LR and LN+H+DSM groups. This should be clarified in the legend. If this p-value comes from comparing HPV cfDNA copies for the (LR, LNM, HM) and (LN+HM, LN+HM+DSM) groups, did the authors carry out post-hoc pairwise comparisons? It would be helpful to include acronyms for these groups in the legend also.

(3) Interpretation of results in Figure 2 and elsewhere: Significant differences detected in Figure 2B could imply potential associations between HPV cfDNA levels (or subtypes) and recurrence/metastasis patterns. Figure 2C shows that there is a difference in cfDNA levels between the groups compared, suggesting an association but this would not necessarily be a direct "correlation". Overall, interpretation of statistical findings would benefit from more precise language throughout the text and overstatement should be avoided.

(4) The authors state that six patients showed cfDNA elevation with clinically progressive disease, yet only three are represented in Figure 3B1 under "Patients whose disease progressed during treatment." What is the expected baseline variability in cfDNA for patients? If we look at data from patients with early-stage cancer would we see similar fluctuations? And does the degree of variability vary for different HPV subtypes? Without understanding the normal fluctuations in cfDNA levels, interpreting these changes as progression indicators may be premature.

(5) It would be helpful if where p-values are given, the test used to derive these values was also stated within parentheses e.g. (P < 0.05, permutation test with Benjamini-Hochberg procedure).

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

Summary:

The study "Monitoring of Cell-free Human Papillomavirus DNA in Metastatic or Recurrent Cervical Cancer: Clinical Significance and Treatment Implications" by Zhuomin Yin and colleagues focuses on the relationship between cell-free HPV (cfHPV) DNA and metastatic or recurrent cervical cancer patients. It expands the application of cfHPV DNA in tracking disease progression and evaluating treatment response in cervical cancer patients. The study is overall well-designed, including appropriate analyses.

Strengths:

The findings provide valuable reference points for monitoring drug efficacy and guiding treatment strategies in patients with recurrent and metastatic cervical cancer. The concordance between HPV cfDNA fluctuations and changes in disease status suggests that cfDNA could play a crucial role in precision oncology, allowing for more timely interventions. As with similar studies, the authors used Droplet Digital PCR to measure cfDNA copy numbers, a technique that offers ultrasensitive nucleic acid detection and absolute quantification, lending credibility to the conclusions.

Weaknesses:

Despite including 28 clinical cases, only 7 involved recurrent cervical cancer, which may not be sufficient to support some of the authors' conclusions fully. Future studies on larger cohorts could solidify HPV cfDNA's role as a standard in the personalized treatment of recurrent cervical cancer patients.

(1) The authors should provide source data for Figures 2, 3, and 4 as supplementary material.

We greatly appreciate your evaluation of our study and fully agree with the limitations you have pointed out. We appreciate your constructive feedback. Based on your suggestions, we have made the following additions to the article. We have realized that the information provided in Figures 2, 3, and 4 is limited. Therefore, we have presented the original data from Figures 2, 3, and 4 in tabular form in Supplementary Table 2.

(2) Description of results in Figure 2: Figure 2 would benefit from clearer annotations regarding HPV virus subtypes. For example, does the color-coding in Figure 2B imply that all samples in the LR subgroup are of type HPV16? If that is the case, is it possible that detection variations are due to differences in subtype detection efficiency rather than cfDNA levels? The authors should clarify these aspects. Annotation of Figure 2B suggests that the p-value comes from comparing the LR and LN + H + DSM groups. This should be clarified in the legend. If this p-value comes from comparing HPV cfDNA copies for the (LR, LNM, HM) and (LN + HM, LN + HM + DSM) groups, did the authors carry out post-hoc pairwise comparisons? It would be helpful to include acronyms for these groups in the legend also.

We fully agree with your point regarding the need for clearer labeling of HPV genotypes in Figures 2B and 2C. If each data point could be color-coded to represent the HPV genotype, Figures 2B and 2C would be clearer and provide more information. However, we must acknowledge that due to the limitations of our current graphing software and our graphical expertise, we were unable to fully represent each HPV genotype in the figures. To address this, we have presented the data in Supplementary Table 2. This table shows the HPV genotype for each patient, the corresponding metastasis patterns, and the baseline HPV copy numbers. We hope this will address the limitation of insufficient information in Figure 2.

The point you raised regarding whether the differences in detection results might stem from variations in subtype detection efficiency rather than cfDNA levels is a valid limitation of this study. Due to the limited sample size, we did not perform subgroup analyses based on different HPV genotypes, which may have introduced bias in the results presented in Figures 2B and 2C. In response, we have added the following clarification in the discussion section (lines 416-422) and addressed this limitation in the limitations section (lines 499-502). Based on your suggestion, we believe that it is essential to expand the sample size and perform subgroup analysis of the baseline copy numbers for each HPV genotype before treatment. We hope to achieve this goal in future studies.

Thank you for your thoughtful comments regarding the statistical analyses in the study. The p-value in Figure 2B comes from the comparison among five groups, using a two-sided Kruskal-Wallis test. Your suggestion to perform post-hoc pairwise comparisons is excellent and has made the data presentation in the article more rigorous. Following your advice, we conducted pairwise comparisons between the groups. We used the Mann-Whitney U test to compare HPV cfDNA copy numbers between two groups. Since the LR group only had one value, it could not be included in the pairwise comparisons. Significant differences were observed in two comparisons: LNM vs. LN + H + DSM (P = 0.006) and HM vs. LN + H + DSM (P = 0.036). No significant differences were found between the other groups: LNM vs. HM (P = 0.768), LNM vs. LN + HM (P = 0.079), HM vs. LN + HM (P = 0.112), and LN + HM vs. LN + H + DSM (P = 0.145), as determined by the Mann-Whitney U test (Figure 2B). (Lines 258-263).

Thank you for your thoughtful suggestion regarding the inclusion of group acronyms in the legends of Figures 2B and 2C. Including the full names corresponding to the abbreviations would indeed enhance clarity. While we attempted to add both acronyms and full names to the figure legend, the full names were too lengthy and impacted the figure's presentation. Therefore, we have provided the full names corresponding to the abbreviations in the figure caption below, to help readers easily understand the abbreviations used in the figure.

(3) Interpretation of results in Figure 2 and elsewhere: Significant differences detected in Figure 2B could imply potential associations between HPV cfDNA levels (or subtypes) and recurrence/metastasis patterns. Figure 2C shows that there is a difference in cfDNA levels between the groups compared, suggesting an association but this would not necessarily be a direct "correlation". Overall, interpretation of statistical findings would benefit from more precise language throughout the text and overstatement should be avoided.

Thank you for your insightful comments regarding the interpretation of results in Figure 2 and elsewhere. We acknowledge that there are several limitations in this study, and the interpretation of the results should be more careful and cautious. Indeed, in the results section, there were issues with inaccurate wording and exaggeration. We have made revisions in the discussion section, which are presented as follows: Preliminary results indicate that baseline HPV cfDNA levels may be linked to recurrence/metastasis patterns, potentially reflecting tumor burden and spread (Lines 411-413). Additionally, we have also made changes in the conclusion section, which are presented as follows: The baseline copy number of HPV cfDNA may be associated with metastatic patterns, thereby reflecting tumor burden and the extent of spread to some extent (Lines 511-513).

(4) The authors state that six patients showed cfDNA elevation with clinically progressive disease, yet only three are represented in Figure 3B1 under "Patients whose disease progressed during treatment." What is the expected baseline variability in cfDNA for patients? If we look at data from patients with early-stage cancer would we see similar fluctuations? And does the degree of variability vary for different HPV subtypes? Without understanding the normal fluctuations in cfDNA levels, interpreting these changes as progression indicators may be premature.

Thank you for your feedback. We appreciate your thorough review and attention to detail. Six cervical squamous cell carcinoma (SCC) patients exhibited elevated HPV cfDNA levels as their clinical condition progressed. In the previous Figures 3A1 and 3A2, we only presented data from three patients, as we initially believed that displaying the cfDNA curves from three patients would offer a clearer view, while including six patients might lead to overlap and reduce clarity. However, this may have caused confusion for readers. Based on your suggestion, we have revised Figure 3A1 to include the cfDNA curves for all six patients who with squamous cell carcinoma who experienced clinical disease progression during treatment (Figure 3A1), along with the corresponding SCC-Ag curves (Figure 3A2).

Thank you for highlighting the issue of baseline variability in HPV cfDNA. This is indeed a limitation of our study, which did not address this aspect. If baseline variability is defined as changes in HPV cfDNA levels measured at different time points before treatment in the same patient, fluctuations at different time points are inevitable and objective. Following your suggestion, we have added a discussion on baseline variability in the limitations section of the manuscript to provide readers with a more objective understanding of our study's findings (Lines 501-502).In future studies, we will incorporate baseline variability into the research design to better understand pre-treatment HPV cfDNA fluctuations and provide support for clinical decision-making.

(5) It would be helpful if where p-values are given, the test used to derive these values was also stated within parentheses e.g. (P < 0.05, permutation test with Benjamini-Hochberg procedure).

Thank you for your valuable suggestions and examples. Following your advice, we have included the statistical test methods used to obtain the p-values in parentheses wherever they appear in the results section. Additionally, we have specified the statistical test methods for the p-values below the figures in the results section.

Reviewer #2 (Public review):

Summary:

The authors conducted a study to evaluate the potential of circulating HPV cell-free DNA (cfDNA) as a biomarker for monitoring recurrent or metastatic HPV+ cervical cancer. They analyzed serum samples from 28 patients, measuring HPV cfDNA levels via digital droplet PCR and comparing these to squamous cell carcinoma antigen (SCC-Ag) levels in 26 SCC patients, while also testing the association between HPV cfDNA levels and clinical outcomes. The main hypothesis that the authors set out to test was whether circulating HPV cfDNA levels correlated with metastatic patterns and/or treatment response in HPV+ CC.

The main claims put forward by the paper are that:

(1) HPV cfDNA was detected in all 28 CC patients enrolled in the study and levels of HPV cfDNA varied over a median 2-month monitoring period.

(2) 'Median baseline' HPV cfDNA varied according to 'metastatic pattern' in individual patients.

(3) Positivity rate for HPV cfDNA was more consistent than SCC-Ag.

(4) In 20 SCC patients monitored longitudinally, concordance with changes in disease status was 90% for HPV cfDNA.

This study highlights HPV cfDNA as a promising biomarker with advantages over SCC-Ag, underscoring its potential for real-time disease surveillance and individualized treatment guidance in HPV-associated cervical cancer.

Strengths:

This study presents valuable insights into HPV+ cervical cancer with potential translational significance for management and guiding therapeutic strategies. The focus on a non-invasive approach is particularly relevant for women's cancers, and the study exemplifies the promising role of HPV cfDNA as a biomarker that could aid personalized treatment strategies.

Weaknesses:

While the authors acknowledge the study's small cohort and variability in sequential sampling protocols as a limitation, several revisions should be made to ensure that (1) the findings are presented in a way that aligns more closely with the data without overstatement and (2) that the statistical support for these findings is made more clear. Specific suggestions are outlined below.

(1) Line 54 in the abstract refers to 'combined multiple-metastasis pattern' but it is not clear what this refers to at this point in the text.

Thank you for your detailed feedback. You are correct that the "combined multi-metastatic pattern" was not adequately explained in the abstract, which may have caused confusion. To address this, we have clarified the definitions of the combined multi-metastatic pattern and single-metastatic pattern in lines 53-55 of the manuscript. Patients with a combined multi-metastatic pattern (lymph node + hematogenous ± diffuse serosal metastasis) exhibited a higher median baseline HPV cfDNA level compared to those with a single-metastasis pattern (local recurrence, lymph node metastasis, or hematogenous metastasis) (P = 0.003).

(2) Line 90 The reference to 'prospective clinical study (NCT03175848) in primary stage IVB CC to investigate the role of radiotherapy (RT) in combination therapy' seems not to be at all relevant at this point in the text. I would limit the description of this study to the methods.

Thank you for your thoughtful and thorough review. Your suggestions are highly relevant. Upon further reflection, we recognized that this sentence was redundant in its original placement. Following your recommendation, we have removed it from this section and moved it to the methods section (Lines 109-111). The revised statement is as follows: "Notably, 19 cases from the primary CC group participated in our prospective clinical study (NCT03175848), focused on stage IVB cervical cancer."

(3) Line 56 refers to HPV cfDNA levels (range 0.3-16.9) but what units?

Thank you for your feedback regarding the manuscript format. While you highlighted this specific issue, we have since identified several other instances of omitted units in parentheses throughout the manuscript. We acknowledge that such formatting oversights can create ambiguity for readers. Following your suggestions, we have corrected all such issues in the manuscript. We greatly appreciate your careful and thorough review.

(4) Lines 247-248 claim that higher baseline HPV cfDNA levels correlated with a more substantial post-chemotherapy decrease. This correlation should be statistically validated, and the p-value should be included.

Thank you for your insightful comments, which highlighted an issue with this sentence. Upon review, I have made the necessary revisions. Since no statistical analysis was conducted and the P-value was not provided, the original sentence was imprecise. Given the small sample size, statistical analysis is not feasible. I have revised the sentence as follows: “For patients in whom systemic cytotoxic chemotherapy was effective, a significant decrease in HPV cfDNA levels could be detected after chemotherapy” (Lines 297-298).

(5) The authors mention that baseline samples were collected "between Day -14 and Day +30 preceding initial treatment." If Day -14 indicates two weeks before treatment, then this would imply some samples were taken up to 30 days post-treatment. This notation should be clarified. To what extent might outliers or more extreme values in Figure 2 driven by variability in how baseline sampling was carried out?

Thank you for your insightful comments. Undoubtedly, this is indeed a major limitation of our study. These factors could lead to a certain degree of bias in the detection data. The primary reason is that the study was conducted during the COVID-19 pandemic, making it sometimes difficult to conduct sampling regularly. In accordance with your suggestion, I have already added this part of the content to the results section of the article (Lines 266-275). We have also included the variation in baseline sampling as a limitation in the discussion section (Lines 497-499). In future studies, we will strive to improve the study design by ensuring baseline samples are collected prior to treatment, thereby enhancing the reliability of statistical and analytical results.

(6) Would be useful to amend Figure 1 to show a subset of patients with SCC and a subset of patients who underwent longitudinal monitoring.

Thank you for your detailed suggestion. Including a subset of pathological types could indeed add more information to Figure 1. However, regarding the pathological types of the patients in this group, we have listed them in Table 1 and Supplementary Table 2. Among the 28 patients, 26 are diagnosed with squamous cell carcinoma, so 92.9% of the patients in this study have squamous cell carcinoma. To avoid making Figure 1 too complex, we decided not to include the pathological type in the figure.

(7) Line 120 "a time point matching or closely following HPV cfDNA sampling" - what is the time range for 'closely following' here? A couple of hours or days after sampling?

Thank you for your detailed feedback. Based on your suggestion, we have revised the sentence as follows:

"For patients with squamous cell CC in the sequential sampling group, concurrent SCC-Ag testing was performed at a time point that matched, or was within 7 days before or after, the HPV cfDNA sampling." (Line 123-125)

(8) Lines 178-190 and lines 179-180 seem to make exactly the same point.

Thank you very much for your careful review. Indeed, these two sentences were repetitive and conveyed the same point. I have removed the previous sentence here (lines 206-207).

(9) In Figure 4, please indicate the number of patients in each group in the legend e.g. HPV16+ (n=x number of patients).

Thank you for your feedback on the details of Figure 4 and the examples provided. We have updated Figure 4 according to your suggestions and included the number of patients in each group in the figure legend.

(10) Lines 322-3 'HPV cfDNA predicted treatment response or disease progression at an earlier time point than imaging assessments' - based on the data available and the numbers of patients, I would argue that this is too bold a claim.

Thank you very much for pointing out this issue. We fully agree with your view. We have modified this sentence as follows: "Secondly, dynamically monitored HPV cfDNA levels appeared to predict treatment response and disease progression. " (Lines 391-392).

Version 2

Abstract

Abstract

Purpose

Monitoring of circulating human papillomavirus (HPV) cell-free DNA (cfDNA) is a minimally invasive approach for surveillance in HPV-associated cancers, particularly cervical cancer. The aim of this study was to monitor circulating HPV cfDNA levels in patients with recurrent or metastatic cervical cancer during treatment and follow-up, assessing the utility of HPV cfDNA as a tumor marker for disease surveillance and in guiding clinical treatment decisions.

Experimental Design

In this prospective pilot observational study, levels of HPV cfDNA in serum samples from 28 patients with recurrent or metastatic HPV-positive cervical cancer were measured via digital droplet polymerase chain reaction. Results for HPV cfDNA levels were matched to clinical outcomes and to serum levels of squamous cell carcinoma antigen (SCC-Ag) to assess the clinical potential of HPV cfDNA as a tumor marker.

Results

HPV cfDNA was detected in all 28 patients (100% detection rate). Notably, median baseline HPV cfDNA levels varied according to the metastatic pattern observed in individual patients (P=0.019). Patients with a combined multi-metastatic pattern (lymph node + hematogenous ± diffuse serosal metastasis) exhibited a higher median baseline HPV cfDNA level compared to those with a single-metastasis pattern (local recurrence, lymph node metastasis, or hematogenous metastasis) (P = 0.003). All participants exhibited changes in HPV cfDNA levels over a median monitoring period of 2 months (range 0.3 to 16.9 months) prior to evaluations for treatment response or disease progression. Among 26 patients initially diagnosed with squamous cell cervical cancer, the positivity rate was 100% for HPV cfDNA and 69.2% for SCC-Ag (P = 0.004, 95% confidence interval, 0-0.391). Among 20 patients longitudinally monitored for squamous cell cervical cancer, the concordance with changes in disease status was 90% for HPV cfDNA and 50% for SCC-Ag (P = 0.014, 95% confidence interval, 0.022-0.621).

Conclusions

HPV cfDNA is a promising tumor marker for HPV-positive cervical cancer. In the context of precision medicine, HPV cfDNA is poised to play an increasingly pivotal role in monitoring treatment efficacy, providing valuable insights into disease progression, and guiding clinical decisions.

Introduction

Cervical cancer (CC) is the fourth most prevalent malignancy in terms of both incidence and mortality among females globally, and it is the primary human papillomavirus (HPV)-associated cancer [1, 2]. Despite advances in CC treatment, challenges persist regarding recurrence and metastasis [3]. The treatment landscape for recurrent or metastatic CC has changed with the emergence of targeted immunotherapy drugs. However, the lack of effective biomarkers hinders the assessment of treatment efficacy and the ability to predict patient outcomes in this setting [4]. There is a pressing need for more effective and minimally invasive biomarkers for serial monitoring of treatment responses and prognosis of patient outcomes.

As a liquid biopsy modality, measurement of HPV cfDNA released from tumor cells into the bloodstream has extensive utility in optimizing various facets of cancer management, including early diagnosis [5], noninvasive genotyping, pretreatment assessment, drug target identification, resistance detection [6, 7], treatment efficacy monitoring, post-treatment follow-up, and relapse prediction [8–10]. HPV is the cause of most CC cases [11]. Viral DNA from high-risk HPV genotypes integrates into host cell genomes, resulting in widespread expression of virus-specific E6/E7 proteins [12]. As HPV-associated cancers release HPV cfDNA into the host’s bloodstream, circulating HPV cfDNA in free or integrated form is an attractive potential biomarker that can be detected in blood [13, 14]. Typically, circulating HPV cfDNA is detected using digital droplet polymerase chain reaction (ddPCR) or next-generation sequencing (NGS) technology [15]. DdPCR allows direct, independent, and absolute quantification of HPV cfDNA in samples, with a DNA detection threshold as low as 1 copy/mL. Results can be obtained within 1 day, therefore ddPCR is relatively cost-effective [16, 17].

Recent advances in targeted therapies and immunotherapies have significantly enhanced the treatment efficacy for recurrent or metastatic CC and have increased overall survival (OS) [4]. Some patients may require maintenance therapy for up to 2 years, and others may survive with their tumor for an extended period. Consequently, it is imperative to closely monitor each patient’s condition and adjust their treatment plans accordingly. However, conventional imaging methods occasionally fail to reflect disease changes in a timely manner [18], while blood biomarkers such as squamous cell carcinoma antigen (SCC-Ag) have limited clinical effectiveness for monitoring [19]. Considering these clinical challenges, we hypothesized that circulating HPV cfDNA levels correlate with metastatic patterns and treatment response in CC. To validate this hypothesis, we prospectively recruited an observational pilot cohort comprising patients with primary stage IVB or recurrent HPV-positive CC and measured HPV cfDNA levels. The primary objective was to examine the correlation between HPV cfDNA copy numbers, disease parameters, and treatment responses. The secondary objective was to compare the predictive value of HPV cfDNA and SCC-Ag in assessing treatment responses.

Methods

Study design

From August 2017 to February 2023, a total of 33 patients with pathologically confirmed primary stage IVB or recurrent HPV-positive CC were enrolled at Zhejiang Cancer Hospital. Five patients were excluded from the analysis, as outlined in Figure 1. The final analysis cohort comprised 28 cases: 21 with primary stage IVB CC and seven with recurrent CC. Notably, 19 cases in the primary CC group participated in the prospective clinical study (NCT03175848) initiated by us, focused on stage IVB cervical cancer. The main eligibility criteria were: (1) Pathologically confirmed diagnosis of CC; (2) Pathological evidence of at least one recurrence or metastatic lesion; (3) PCR positivity for a high-risk HPV subtype in pretreatment exfoliated cervical cells or a serum sample. All patients consented to the study protocol, including collection of blood samples throughout the study. Patients received chemotherapy with or without immunotherapy, targeted therapy, or radiotherapy (RT). Exclusion criteria: (1) Non-HPV-related cervical cancer; (2) No measurable lesions; (3) Synchronous or multiple cancers; (4) Patient refusal to provide blood samples; (5) Patient refusal to undergo regular follow-up and periodic imaging evaluations. For patients with primary stage IVB CC who had a single sample collected, the sample was obtained at treatment initiation. For the group undergoing longitudinal sampling, patients with primary stage IVB CC had three to five blood samples collected at treatment initiation, mid-treatment, and during follow-up, while patients with recurrent CC had three blood samples collected during treatment, starting from enrollment. For all serum specimens, HPV cfDNA was quantified using ddPCR. For patients with squamous cell CC in the sequential sampling group, concurrent SCC-Ag testing was performed at a time point that matched, or was within 7 days before or after, the HPV cfDNA sampling.

CONSORT flow diagram showing the enrollment of an observational cohort of patients diagnosed with recurrent or metastatic HPV-positive cervical cancer undergoing serum HPV cell-free DNA surveillance.

The 2018 International Federation of Gynecology and Obstetrics (FIGO) staging criteria were applied. Each patient underwent routine imaging assessments, and treatment efficacy was evaluated according to Response Evaluation Criteria in Solid Tumors (RECIST version 1.1). During treatment, imaging evaluations are typically conducted every two cycles of chemotherapy. Imaging assessments are also conducted before and after radiotherapy. For patients receiving immunotherapy or targeted therapy maintenance, assessments should be performed at least every 3 months. In follow-up, imaging is recommended every 3 months for the first 2 years and every 6 months from years 3 to 5, though this schedule may be adjusted based on individual patient needs. We defined lymph node metastasis as metastasis in para-aortic lymph nodes or distant lymph nodes, such as supraclavicular, inguinal, or mediastinal nodes. We classified patterns of CC recurrence or metastasis into five groups: local recurrence (LR); lymph node metastasis (LNM); hematogenous metastasis (HM); lymph node + hematogenous metastasis (LN + HM); and lymph node + hematogenous + diffuse serosal metastasis (LN + H + DSM). DSM encompasses metastases to the peritoneal, pleural, or pericardial regions. Prior to initiating treatment with immune checkpoint inhibitors (ICIs), tumor tissue specimens should be tested for PD-L1 expression using immunohistochemistry (IHC), provided that tumor samples are available. The PD-L1 IHC 22C3 pharmDx assay is used for testing, and PD-L1 expression in cervical cancer is quantified by the Combined Positive Score (CPS). The Ethics Committee of Zhejiang Cancer Hospital approved the study.

HPV diagnostics and typing

HPV genotypes were determined via routine PCR for exfoliated cervical cells or serum samples collected before treatment. Patients were excluded from the study if both samples tested negative. A commercial PCR testing kit (Hybribio HPV Genotyping Detection Kit), designed primarily for qualitative genotyping of HPV DNA extracted from exfoliated cervical cells, was routinely used. HPV genotyping detection was performed using PCR combined with the hybridization method. This kit can detect 21 HPV genotypes, including 8 low-risk genotypes and 13 high-risk genotypes. The 13 high-risk genotypes include: HPV16, HPV18, HPV31, HPV33, HPV35, HPV39, HPV45, HPV51, HPV52, HPV56, HPV58, HPV59, and HPV68.

The following steps outline the procedure: A disposable cervical exfoliation brush is placed at the cervical opening and rotated gently clockwise five times. The brush is then transferred to a sample tube containing 10 mL of cell preservation solution. The tube is stored at 4℃ and tested within one week. The sample is centrifuged at 3,000 rpm for 15 minutes, and the supernatant is discarded. Next, 200 µL of cell preservation solution is added to resuspend the cells. DNA is extracted using the QIAamp DNA Mini Kit (Qiagen, Hilden, Germany), and PCR amplification is performed according to the manufacturer’s instructions (total reaction volume: 20 µL/sample), followed by setting baseline and threshold values for result interpretation. One patient tested positive for HPV type 6 via PCR on cervical exfoliated cells, with no high-risk HPV genotypes detected. IHC for P16 on CC tissue was negative, leading to the classification of the tumor as non-HPV-associated and exclusion from the study.

Samples

Serum was extracted from peripheral whole blood samples at the Radiobiology Laboratory of Zhejiang Cancer Hospital. A 5-mL aliquot of whole blood was collected into a yellow-top blood collection tube and allowed to clot at room temperature for 30 minutes. After centrifugation at 2,000 × g for 10 minutes in a refrigerated centrifuge, the serum was carefully transferred into polypropylene tubes in 1-mL aliquots and stored at −80°C. Frozen serum samples were subsequently transported to the Oncology Research Institute of Zhejiang Cancer Hospital for DNA extraction.

Procedure for ddPCR analysis

Before analysis, serum samples were thawed and centrifuged at 2,000 × g at 4°C for 10 minutes for DNA extraction. In accordance with the manufacturer’s protocol, HPV cfDNA was isolated from 2 mL of serum using a QIAamp Circulating Nucleic Acid Extraction Kit (Qiagen, Hilden, Germany). The DNA was eluted twice through a column for purification, resulting in 60 μL of eluate that was stored at −80°C until analysis. Primers and probes for ddPCR detection were designed based on E7 gene sequences of the target HPV genotypes to generate amplicons of varying length (Supplementary Table 1). Each ddPCR reaction used 30 μL of DNA template. According to the manufacturer’s instructions for the QIAcuity QX-200 ddPCR platform (Qiagen), ddPCR reactions consisted of 40 µL of reaction mixture per well that included the primers, probes, and template. The reactions were amplified in QIAcuity 26,000 24-well Nanoplates (Qiagen) under the following conditions: initial enzyme activation at 95°C for 2 min, followed by 50 cycles of denaturation at 95°C for 15 seconds and annealing at 60°C for 30 seconds. The exposure time for imaging of partitions was 400 ms for fluorescein amidite (FAM) and 300 ms for 2’-chloro-7’-phenyl-1,4-dichloro-6-carboxyfluorescein (VIC). Data analysis was performed using QIAcuity software version 2.1.7 (Qiagen) to quantify HPV copy numbers.

HPV cfDNA monitoring protocol

Participants were enrolled on a rolling basis, and the first serum sample collected was regarded as the baseline sample. A baseline sample was categorized as a treatment initiation sample if it was collected between Day−14 and Day+30 preceding initial treatment in patients with primary stage IVB CC, or before treatment for relapse or disease progression in patients with recurrent CC.

For patients with sequential samples, the time at which the initial blood sample was collected was designated as time 0. Up to four additional Blood samples were collected at intervals ranging from 8 to 962 days, with a median of 73 days and a mean of 128 days, and blood collection continued for up to 1513 days. Blood sample collection was coordinated with patient treatment and follow-up times to the greatest extent possible for measurement of HPV cfDNA levels (copies/mL). Serum SCC-Ag levels are routinely assessed multiple times before, during, and after treatment for patients with squamous cell CC at our hospital using an Abbott Architect instrument (Abbott Laboratories, Abbott Park, IL, USA). To ensure consistency in the analysis, we selected the SCC-Ag values measured simultaneously with the HPV serum samples, or chose the most recent available SCC-Ag values within 7 days. SCC-Ag levels < 1.5 ng/mL were classified as normal, while levels >= 1.5 ng/mL were classified as elevated.

The upper limit of detection was 70 ng/mL. For statistical purposes, clinical test results exceeding 70 ng/mL were treated as 70 ng/mL.

Following the literature [20, 21], serum samples were deemed HPV-positive if at least three droplets containing HPV amplicons were identified. Samples with fewer than three droplets containing HPV amplicons or no amplicons detected were categorized as HPV-negative. Serum HPV cfDNA levels were quantified as copies/mL.

Statistical analyses

Statistical analyses were performed using GraphPad Prism 9. The Mann-Whitney U test was used to assess the difference in the number of viral DNA copies between two groups or two metastasis patterns.

Kendall’s τ correlation test was used to determine the coefficient of correlation between two factors. Rate comparisons were conducted using Fisher’s exact test. Kaplan-Meier survival analysis was performed to calculate the hazard ratio with a 95% confidence interval (CI), using the Cox model.

Overall survival (OS) was defined as the time from diagnosis until death from any cause or the last follow-up date (December 31, 2023). All P-values reported are two-tailed, with statistical significance defined as P < 0.05.

Results

Patient characteristics

Our study included 28 patients diagnosed with HPV-positive advanced CC treated at our hospital, comprising 21 cases (75%) with primary stage IVB CC and seven cases (25%) with recurrence and metastasis after treatment. The clinical characteristics of the patients are listed in Table 1. The median age at diagnosis was 52 years (range 34–67 years). In total, 76 serum samples were obtained from the cohort, consisting of 69 longitudinal samples from 21 patients (3–5 samples per patient) and single-time samples from seven patients at treatment initiation. Of the baseline samples, 25 (89%) were obtained at treatment initiation. HPV cfDNA was quantified in all serum samples via ddPCR. Sixteen patients received ICIs, of whom 12 also received concurrent targeted therapy, while four received ICIs alone. Three patients with stage IVB disease were treated with Paclitaxel/Cisplatin (TP) + Bevacizumab (Bev) + ICIs as first-line therapy, and 13 patients received ICIs after disease progression. The specific treatment regimens and HPV cfDNA copy numbers for each patient are provided in Supplementary Table 2.

Validity of the test for HPV cfDNA detection and genotyping

To validate the consistency of HPV genotyping between serum, and exfoliated cervical cells from the same patient, we conducted conventional PCR genotyping of HPV cfDNA in baseline serum. The analysis revealed that only 12/28 patients (42.9%) tested positive, however, the genotyping outcomes were entirely consistent with those obtained from matched cervical exfoliated cells.

Subsequently, the 28 baseline serum samples underwent qualitative and quantitative assessment using ddPCR. The results revealed that all 28 patients (100%) tested positive for HPV cfDNA, with HPV typing showing complete concordance (100%) with PCR results from matched cervical exfoliated cells. HPV genotyping revealed that HPV16, HPV58, HPV18, and other genotypes (HPV31, HPV66, HPV33) accounted for 72.4%, 10.3%, 6.9%, and 10.3%, respectively (Figure 2A).

(A) Serum HPV cfDNA copy number for six HPV genotypes at baseline.

(B) Baseline HPV cfDNA copy numbers for five recurrence/metastasis subgroups. Statistical significance was determined using a two-sided Kruskal-Wallis test. Abbreviations: LR, local recurrence; LNM, lymph node metastasis; HM, hematogenous metastasis; LN + HM, lymph node + hematogenous metastasis; LN + H + DSM, lymph node + hematogenous + diffuse serosal metastasis. (C) Relationship between HPV cfDNA copy number and metastatic pattern. Statistical significance was determined using a two-sided Mann-Whitney U test. All plots show the median and interquartile range on a log10 scale.

Correlation between tumor metastasis pattern and baseline HPV cfDNA

Analysis for the study cohort revealed an association between tumor metastatic pattern and baseline HPV cfDNA levels. According to their metastatic status at baseline, the patients were categorized into two groups: the Single-Metastasis Pattern (SMP) group (LR, LNM, or HM); and the Multi-Metastatic Pattern (MMP) group (LN + H ± DSM). The baseline copy number significantly differed among the five recurrence/metastasis pattern groups (P = 0.019, with Kruskal-Wallis test) and tended to increase with the degree of metastasis (Figure 2B). In the subsequent group comparisons, the LR group was excluded due to containing only a single value, leaving four groups for comparison. Significant differences were observed in two comparisons: LNM vs. LN + H + DSM (P = 0.006) and HM vs. LN + H + DSM (P = 0.036). No significant differences were found between the other groups: LNM vs. HM (P = 0.768), LNM vs. LN + HM (P = 0.079), HM vs. LN + HM (P = 0.112), and LN + HM vs. LN + H + DSM (P = 0.145), as determined by the Mann-Whitney U test (Figure 2B). The median baseline HPV cfDNA copy number was significantly higher in the MMP group than in the SMP group (P=0.003, with Mann-Whitney U test, Figure 2C).

Baseline samples for patients with primary stage IVB cervical cancer were collected between Day-14 to +30 relative to the initial treatment. Some patients provided pre-treatment samples, while others provided post-treatment samples, with the latter potentially influenced by the treatment itself. To assess whether there was a significant difference in baseline HPV cfDNA levels before and after treatment (within 30 days), a statistical analysis was performed. To minimize the impact of HPV genotypes, we focused on the baseline HPV cfDNA values of 14 HPV16-positive patients with primary stage IVB CC (6 pre-treatment and 8 post-treatment samples). The median HPV cfDNA value for pre-treatment samples was 8.01 × 10L copies/mL (range 7.5 × 10³–9.7 × 10L copies/mL), while the median for post-treatment samples was 3.5 × 10L copies/mL (range 4.7 × 10²–7.4 × 10L copies/mL). “The difference between the two groups was not statistically significant (P = 0.414, with Mann-Whitney U test).

HPV cfDNA as a predictor of treatment response or failure

We followed the enrolled patients for an average of 24.1 months (range 2.1 to 77.8 months). Figure 3 shows temporal changes in serum HPV cfDNA levels. In all patients with longitudinal testing, changes in HPV cfDNA levels occurred at a median of 2 months (range 0.3–16.9 months) before imaging confirmation of a treatment response or disease progression. Six patients with squamous cell carcinoma who experienced clinical disease progression during treatment (as per RECIST) exhibited an elevation in HPV cfDNA copy numbers before imaging-confirmed progression.The median time from detection of elevated plasma HPV cfDNA to imaging confirmation of disease progression was 4.2 months (range 1.9–16.9 months; Figure 3A1). Similarly, we observed a consistent decrease in HPV cfDNA copy numbers in 16 patients before imaging confirmed a treatment response. The median time from detection of a decrease in HPV cfDNA to imaging confirmation of disease regression was 1.2 months (range 0.3–2.8 months; Figure 3B1-D1).

Paired plots for serum HPV cfDNA (copies/mL) and squamous cell carcinoma antigen (SCC-Ag) (ng/mL) levels were measured longitudinally in patients with metastatic or recurrent HPV-positive cervical cancer (each patient is represented by a line of the same color). HPV cfDNA levels in serum were scaled using log10.

(A1, A2) Patients whose disease progressed during treatment. (B1, B2) Selected patients who showed a response to cytotoxic chemotherapy. (C1, C2) Response to immune therapy in a patient with stage IVB cervical cancer positive for HPV genotypes 33 and 16, treated with paclitaxel + cisplatin, concurrent chemoradiotherapy, brachytherapy, and bevacizumab. (D1, D2) Selected patients with a response to radiotherapy. (E1, E2) HPV cfDNA levels were significantly higher than normal at some of the longitudinal time points in five patients, but SCC-Ag levels were within the normal range (< 1.5 ng/mL) at all time points. (F1, F2) HPV cfDNA levels for all patients with longitudinal samples (n = 21) and matched SCC-Ag levels for the patients with squamous cell cancer (n = 20). Each colored line corresponds to one patient, except in C. Horizontal dashed lines indicate the days between plasma HPV cfDNA levels suggestive of a response or progression (rise or fall in levels) and imaging-confirmed changes in disease status. The associated number of days (D) is listed adjacent to the horizontal dashed line for comparison. It should be noted that not all HPV cfDNA data points are plotted for each patient.

For patients in whom systemic cytotoxic chemotherapy was effective, a significant decrease in HPV cfDNA levels could be detected after chemotherapy. For two patients with high HPV cfDNA levels of 3.1 × 106 copies/mL and 1.7 × 106 copies/mL at baseline, the HPV cfDNA copy number decreased significantly following one or two cycles of paclitaxel + cisplatin (TP) chemotherapy (median 98.2%, range 96.7–99.5%). The rate of decline in viral load slowed after the subsequent cycle of TP chemotherapy, with a median decline of 51.3% (range 39–63.5%). In addition, a patient with baseline HPV cfDNA of 2.8 × 104 copies/mL experienced a 45% reduction in copy number following one cycle of TP chemotherapy (Figure 3B1).

We observed that changes in HPV cfDNA levels may indicate a response to combined immunotherapy and targeted therapy. One patient diagnosed with primary stage IVB CC and multiple metastases (LN + H + DSM) tested positive for HPV33 and HPV16 in exfoliated cervical cells and serum. The patient received four cycles of TP and concurrent chemoradiotherapy plus brachytherapy for the pelvic-abdominal primary focus, followed by two additional TP cycles. The HPV33 viral load decreased from a pretreatment level of 1.5 × 107 copies/mL to 6.7 × 104 copies/mL after treatment; however, the HPV16 viral load increased from 7.7 × 102 to 1.2 × 104 copies/mL. Subsequent imaging evaluation after 58 days indicated the emergence of new foci in the lungs, prompting adjustments to the patient’s treatment plan. Guided by the immunohistochemical presence of PD-L1-positive cells (CPS = 10) in the primary cervical lesion, the patient received one cycle of TP + Bev + ICIs treatment. This regimen led to a swift decrease in viral levels of both HPV genotypes after 36 days, with HPV 33 reaching undetectable levels and HPV 16 decreasing to 7.8 × 102 copies/mL (Figure 3C1). Three patients with stage IVB disease received TP + Bev + ICIs as first-line therapy. Two patients achieved complete remission (CR) (CPS = 70 and CPS = 5), while one experienced disease progression (CPS = 1). HPV cfDNA levels closely correlated with the clinical outcomes of these patients, demonstrating a sensitive and effective response (Supplementary Table 2).

Serum HPV cfDNA levels also changed in response to RT. A transient rise (20.8-fold) in HPV cfDNA copy number was observed in a patient with stage IVB CC who experienced a 2-week interruption of RT due to grade IV thrombocytopenia. Following Stereotactic Body Radiotherapy (SBRT) to lung metastases, a patient with pulmonary oligometastases experienced a 29% decrease in HPV cfDNA copy number. Another patient, diagnosed with stage IVB CC and pelvic bone metastases, who underwent radical concurrent chemoradiotherapy targeting both the primary and metastatic lesions exhibited a notable reduction in HPV cfDNA levels (Fig. 3D1).

Correlation between HPV cfDNA and SCC-Ag

There were 26 patients with squamous cell CC in the study cohort. All 26 (100%) had elevated serum HPV cfDNA at baseline, but only 18/26 patients (69.2%) had elevated SCC-Ag at baseline (P = 0.004, 95% CI, 0–0.391, with Fisher’s exact test). Among 72 serum samples from patients with squamous cell CC, the median HPV cfDNA level was 1.7×104 copies/mL (range 0–1.4×107 copies/mL) and the median SCC-Ag level was 2.6 ng/mL (range 0.4–70 ng/mL). There was no significant correlation between SCC-Ag and HPV cfDNA levels (R2 = 0.034, P = 0.120, with Kendall’s τ correlation test). For patients with squamous cell CC who had longitudinal monitoring (n = 20), the concordance with disease change was 90% for HPV cfDNA and 50% for SCC-Ag (P = 0.014, 95% CI, 0.022–0.621, with Fisher’s exact test). Comparison of matched serum HPV cfDNA and SCC-Ag levels for patients with squamous cell CC (Figure 3A-D) revealed that HPV cfDNA exhibited dynamic fluctuations, while serum SCC-Ag levels in the majority of patients rapidly decreased to near or below the normal range (< 1.5 ng/mL) following initiation of treatment. During the course of treatment, SCC-Ag levels remained within the normal range (< 1.5 ng/mL) at all time points in five patients, but matched serum HPV cfDNA showed fluctuating changes above normal values at some time points.(Figure 3E1, E2).

Correlation between HPV cfDNA and survival

The 5-year OS rate for the entire cohort was 42.3%, with median OS of 52.1 months at median follow-up of 42.3 months (range 10.2–88.5 months). As of December 31, 2023, there were 12 patient deaths and 20 disease progression events. Analysis of survival by HPV genotypes revealed that the difference in OS between the HPV16+ group and the non-HPV16+ group was not statistically significant (P = 0.052, with log-rank test; Figure 4A). Univariate and multivariate analyses were conducted to evaluate the association between clinicopathologic factors and patient OS, with no factors identified as significantly impacting OS. The results of the univariate analysis are shown in Supplementary Table 3. Correlation analysis of baseline HPV cfDNA copy number with mortality and OS outcomes revealed no significant association (R²= –0.111, P = 0.486; R²= –0.037, P = 0.782, with Kendall’s τ correlation test).

(A) Overall survival for patients stratified by HPV subtype (HPV 16+ versus non-HPV16+).

(B) Overall survival for patients stratified by baseline serum HPV cfDNA copy number, using a cutoff of 3.9×104 copies/mL. P-values were calculated using a two-sided log-rank test.

Patients with primary stage IVB CC (n = 21) were stratified by baseline median HPV cfDNA level using 3.9×104 copies/mL as the dichotomization threshold. The difference in OS between the groups with ≥ 3.9×104 copies/mL and < 3.9×104 copies/mL at baseline was not statistically significant (P = 0.111, with log-rank test; Figure 4B). Survival analysis of 21 consecutively monitored patients, categorized by the trend in HPV cfDNA levels, revealed two groups: 12 patients with decreasing levels and 9 with increasing levels. The OS difference between these groups was not significant (P = 0.866, with log-rank test). Additionally, patients were divided into two groups based on whether HPV cfDNA levels decreased to normal: three patients with normalization and 18 without. The OS difference between these groups was also not significant (P = 0.590, with log-rank test).

Discussion

We conducted a prospective pilot observational study in patients with metastatic or recurrent CC to analyze ddPCR HPV cfDNA results in relation to SCC-Ag levels, clinical treatment responses, and prognosis. The study findings confirm the significant clinical potential of dynamic HPV cfDNA surveillance for CC. Firstly, we observed a correlation between baseline HPV cfDNA copy number and recurrence/metastasis patterns. Secondly, dynamically monitored HPV cfDNA levels appeared to predict treatment response and disease progression. Finally, in monitoring HPV-associated recurrent and metastatic cervical cancer, HPV cfDNA may offer advantages over SCC-Ag.

Our positivity rate for HPV cfDNA was 42.9% (12/28) when using PCR and 100% (28/28) when using ddPCR. We found 100% agreement in HPV typing results between exfoliated cervical cells and serum samples. These findings showed the high sensitivity and specificity of HPV cfDNA detection via ddPCR. A meta-analysis comparing the accuracy of different methods for HPV cfDNA detection in HPV-positive tumors revealed that NGS outperformed ddPCR and quantitative PCR in terms of sensitivity, while specificity remained consistent across all three methods [15]. We found that HPV cfDNA positivity in HPV-positive CC correlated positively with tumor stage, tumor load, and lymph node status. Consistent with our findings, a recent study demonstrated a serum HPV cfDNA positivity rate of 100% in HPV-positive metastatic CC [22]. Consequently, HPV cfDNA appears to be an ideal serum tumor marker for HPV-positive metastatic or recurrent CC, given the highly sensitive and specific detection methods available.

Studies have shown that HPV cfDNA levels correlate with disease stage, tumor size, tumor load, and lymph node status [23–25]. Accurate quantification of tumor load can be challenging, particularly in settings involving diffuse serosal metastases. Therefore, we classified recurrence/metastasis patterns into five categories and observed a significant difference in median baseline HPV cfDNA copy number among these (P = 0.019). Patients were categorized into SMP and MMP groups according to their recurrence/metastasis status at baseline. The MMP group had a higher median HPV cfDNA copy number at baseline than the SMP group (P = 0.003). Preliminary results indicate that baseline HPV cfDNA levels may be linked to recurrence/metastasis patterns, potentially reflecting tumor burden and spread. A study by Mittelstadt et al. involving 35 patients with CC also revealed correlation between HPV cfDNA levels and tumor load and spread [24]. Patients with multiple metastases are likely to have a higher tumor load with greater shedding of HPV-containing DNA fragments, which can enter the blood circulation via several pathways, resulting in higher HPV cfDNA levels. However, this analysis has several limitations. It did not account for variations in cfDNA detection across HPV genotypes, which may arise from differences in detection rates or expression levels. Subgroup analysis by HPV genotype was not feasible due to the small sample size. These inherent differences may introduce bias, potentially acting as a confounding factor. Future studies will expand the sample size and include subgroup analyses by HPV genotype to better elucidate the relationship between HPV cfDNA copy number and recurrence/metastasis patterns.

In our study, changes in HPV cfDNA levels frequently preceded confirmation of disease changes on imaging scans. Several studies demonstrated that changes in HPV cfDNA copy number are associated with response to therapy for HPV-positive tumors [22, 24, 25]. We found that the median time from the onset of a change in HPV cfDNA copy number to imaging confirmation of a treatment response or disease progression was 2 months (range 0.3–16.9 months). Another study reported analogous findings in HPV-associated oropharyngeal cancer: changes in HPV cfDNA copy number were observed at a median of 16 days (range 12–38 days) before imaging confirmation of treatment response or disease progression in all 22 patients enrolled [26]. These findings suggest that HPV cfDNA can serve as a sensitive marker and a valuable clinical indicator. In addition, our results indicate that HPV cfDNA can serve as a tool for monitoring the effectiveness of RT. Among patients receiving effective RT, HPV cfDNA levels gradually declined throughout the treatment; however, interruptions in RT could lead to transient increases in HPV cfDNA levels. Moreover, our results demonstrate that the combination of immunotherapy and targeted therapy, along with cytotoxic chemotherapy, enhances tumor cell death and clearance of HPV cfDNA in vivo, surpassing the effects of cytotoxic chemotherapy alone. Our findings suggest that HPV cfDNA holds promise as a tool for evaluating the effectiveness of novel therapies such as immunotherapy and targeted therapy and informing subsequent maintenance treatment strategies for patients.

While SCC-Ag is acknowledged as a serum tumor marker for squamous cell CC [27–29], its clinical sensitivity and specificity for monitoring treatment responses are limited [30, 31]. We compared HPV cfDNA and corresponding SCC-Ag levels in 72 serum samples and found no correlation between the two data sets (R2 = 0.03, P = 0.11). HPV cfDNA exhibited several advantages as a serum tumor marker in our study cohort. First, the positivity rate was 100% for HPV cfDNA versus 69.2% for SCC-Ag at baseline (P = 0.004, 95% CI, 0–0.391). Second, changes in HPV cfDNA copy number showed greater concordance with disease progression in the group of 20 patients with squamous cell CC with longitudinal monitoring, with a concordance rate of 90% versus 50% for SCC-Ag (P = 0.014, 95% CI, 0.022–0.621). Finally, HPV cfDNA provides more comprehensive information for dynamic monitoring in comparison to SCC-Ag, as SCC-Ag levels consistently remained within the normal range in some patients. Thus, HPV cfDNA may be a more useful serum marker than SCC-Ag in patients with metastatic or recurrent CC. The lack of correlation between SCC-Ag and HPV cfDNA levels can be attributed to several factors. Firstly, the mechanisms underlying SCC-Ag and HPV cfDNA production differ, leading to distinct treatment responses and dynamic trends. Secondly, the detection methods vary: HPV cfDNA is quantitatively measured by ddPCR, an exponential method, while SCC-Ag is quantified by counting, resulting in significant differences in measurement outcomes. Thirdly, the positivity rates for HPV cfDNA and SCC-Ag differ in this study, further weakening their correlation. In clinical practice, a panel of serum tumor markers is commonly used to monitor baseline conditions prior to treatment. In cases where SCC-Ag is negative, other serum markers may be positive, aiding in the assessment of treatment efficacy and disease progression. Future research into the differing treatment responses of HPV cfDNA and SCC-Ag may help optimize the use of cervical cancer serum biomarkers and their combination.

The 5-year OS rate for the study cohort was 42.3% and the median OS time was 52.1 months. This good outcome was due to active systemic treatment (chemotherapy + immunotherapy) and local treatment (RT to primary and metastatic foci). Therefore, noninvasive dynamic monitoring using serum tumor markers is important for this population. Our data showed that correlation between OS and HPV genotypes (HPV16+ vs non-HPV16+) did not reach statistical significance (P = 0.052). However, we observed a trend towards better prognosis for patients with HPV16+ than for patients with positivity for other HPV genotypes, and further expansion of the sample size may yield positive results. In the analysis of baseline copy number in relation to prognosis, the difference in OS between the groups with baseline HPV cfDNA levels ≥ 3.9×104 and < 3.9×104 copies/mL was not significant (P = 0.111). Previous studies reported a positive correlation between high baseline HPV cfDNA levels and poor prognosis in oropharyngeal cancer [32–34].

Univariate and multivariate analyses, along with correlation analyses, were performed to evaluate factors influencing overall survival (OS) and assess the potential association between HPV cfDNA copy number and patient prognosis. However, the results were not statistically significant. Several factors may have contributed to this, including the small sample size, limited blood samples per patient, and considerable heterogeneity among patients with recurrent metastatic cervical cancer. Additionally, variables such as metastatic lesion status and treatment modalities (e.g., immunotherapy, targeted therapy, or radiotherapy) could have influenced the findings. Therefore, larger studies with extended follow-up, more frequent blood sample collection, and comprehensive analyses are needed to better elucidate the relationship between HPV cfDNA copy number and OS in cervical cancer.

The results of landmark clinical trials, including GOG240, Keynote158, and Keynote826, have integrated ICIs and targeted therapy into the clinical management of recurrent or metastatic CC, significantly improving patient survival [35–37]. This shift was reflected in our enrollment process, where 16 patients received ICIs. Early-phase patients were treated with conventional chemotherapy regimens. After the Keynote-158 study, 13 patients received ICIs upon disease progression, provided that their PD-L1 CPS ≥ 1. Following the Keynote-826 study, three patients with stage IVB disease (LN + HM) were treated with TP + Bev + ICIs as first-line therapy. Two patients (CPS = 70 and CPS = 5) achieved CR, and one (CPS = 1) had remission lasting 13.5 months despite progression. These results suggest that adding ICIs to first-line therapy significantly enhances efficacy in stage IVB cervical cancer. As illustrated in Figure 3, dynamic changes in HPV cfDNA levels in patient C indicate a substantial reduction after one cycle of TP + Bev + ICIs compared to TP chemotherapy and radiotherapy. When other serum markers, like SCC-Ag, were uninformative, HPV cfDNA served as a valuable biomarker, suggesting the potential effectiveness of combining Bev and ICIs. These findings underscore the utility of HPV cfDNA in monitoring the efficacy of ICIs and targeted therapies in HPV-associated cancers.

Our study has several limitations, including a small sample size and a heterogeneous sequential sampling protocol, which introduced variability in baseline blood sampling timing, sample quantities, and intervals between samplings during treatment. Additionally, the heterogeneity of cfDNA levels across different HPV genotypes (due to differences in expression levels or detection efficiencies) and baseline variability of HPV cfDNA within individual patients may also contribute to potential biases in the results. We are currently conducting a prospective study of stage IVB CC. Owing to the absence of literature support for HPV cfDNA sampling protocols, we designed this initial exploratory study with a small sample size to clarify the value of HPV cfDNA monitoring and explore various sampling times and intervals. Our study results suggest that monitoring of HPV cfDNA is valuable before, during, and after treatment. Assessment of HPV cfDNA levels in every chemotherapy cycle (monthly) during treatment and every 3–6 months during follow-up may be a reasonable approach.

Conclusions

Our prospective study suggests that HPV cfDNA, with its high sensitivity and specificity, holds promise as a biomarker for monitoring treatment response and facilitating long-term follow-up in patients with recurrent or metastatic HPV-associated cervical cancer. The baseline copy number of HPV cfDNA may be associated with metastatic patterns, thereby reflecting tumor burden and the extent of spread to some extent. As a serum tumor marker, HPV cfDNA may outperform SCC-Ag in tracking disease dynamics and enabling timely assessment of treatment responses to chemotherapy, radiation, immunotherapy, and targeted therapies. These preliminary findings highlight HPV cfDNA’s potential for monitoring treatment efficacy and predicting disease progression and recurrence in HPV-associated cancers. However, further validation through large-scale prospective trials is needed.